Mental Rotation as Bayesian Quadrature

### 2013-10-15 man/nips-2013-bayesian-optimization.tex
Remove extra space after 'Fig.'

Consider the images on the left in Fig.
\ref{fig:mental-rotation}. In each panel, are the two depicted objects
identical (except for a rotation), or distinct? When presented with
this mental rotation task, people default to a strategy in which they
visualize one object rotating until it is congruent with the other
\cite{Shepard1971}. There is strong evidence for such mental
simulation: we can imagine three-dimensional objects in our minds and
manipulate them, to a certain extent, as if they were real
\cite{Kosslyn:2009tj}.  However, the use of mental simulation is
predicated on determining appropriate parameters to give the
simulation, analogous to determining exactly what computation should
be passed to a graphics processor.  In the case of the classic mental
rotation task, we might ask: How do people know which way to rotate
the object?  When should one stop rotating and accept the hypothesis
that the objects are different?

@@

We begin by first analyzing mental rotation at Marr's
\textit{computational} level \cite{Marr:1983to}: what is the problem
to be solved?  Formally, people are presented with two images, $X_a$
and $X_b$, which are the coordinates of the vertices of 2D shapes
(e.g., Fig.  \ref{fig:stimuli}). Participants must determine whether
$X_a$ and $X_b$ depict the same shape, i.e., whether $\exists
R\textrm{ s.t. } X_b=RX_a$, where $R$ is a rotation matrix. We can
formulate the judgment of whether $X_a$ and $X_b$ have the same
origins by deciding about two hypotheses, $h_0$: $\forall R\ X_b\neq
RX_a$ and $h_1$: $\exists R\textrm{ s.t. } X_b=RX_a$.  To compare the
hypotheses, we need to compute the posterior for each: $p(h\ \vert\
X_a, X_b)\propto p(X_a, X_b\ \vert\ h)p(h)$. Assuming the hypotheses
are equally likely \textit{a priori}, the prior term $p(h)$ will
cancel out when comparing $h_0$ and $h_1$, thus allowing us to focus
on the likelihoods, which are $p(X_a,\ X_b\ \vert \ h_0)=p(X_a)p(X_b)$
for $h_0$ and $p(X_a,\ X_b\ \vert \ h_1)=\int_R p(X_a) p(X_b\vert
X_a,R) p(R)\ \mathrm{d}R$ for $h_1$. From these likelihoods, we
compute the ratio $\ell$ which is given by $\ell=\left(\int_R p(X_b\
  \vert\ X_a, R)p(R)\ \mathrm{d}R\right) /\ p(X_b)$. If $\ell<1$, then
$h_0$ is the more likely hypothesis. If $\ell>1$, then $h_1$ is the
more likely hypothesis.

@@

We define the prior probability of shape $X$ to be
$p(X)=n!\left(\frac{1}{2\pi}\right)^n$ according to a generative
procedure.\footnote{A set of $n$ vertices could be chosen in any of
  $n!$ different ways, and each vertex is located at a random angle
  (between 0 and $2\pi$) and radius (between 0 and 1).} This gives us
the denominator of $\ell$. Computing the numerator is more difficult,
as we we cannot compute $p(X_b\vert X_a, R)$ directly. Instead, we
introduce a new variable $X_R$ denoting a mental image, which
approximates $RX_a$. The $X_R$ are generated by repeated application
of a function $\tau$, i.e. $X_R=RX_a=\tau(X_{R-r},
r)=\tau(\tau(X_{R-2r}, r), r)=\ldots{}=\tau^{(\frac{R}{r})}(X_a, r)$,
where $r$ is a small angle, and $\tau^{(i)}$ indicates $i$ recursive
applications of $\tau$. Using this sequential function, we get:
\begin{align}
  &p(X_a, X_b\ \vert \ h_1)=\int_R \int_{X} p(X_b\vert X) p(X\vert X_a, R)p(X_a)p(R)\ \mathrm{d}X\ \mathrm{d}R \nonumber \\
  &= \int_R \int_X p(X_b\vert X)\delta(\tau^{(\frac{R}{r})}(X_a, r)-X)p(X_a)p(R)\ \mathrm{d}X\ \mathrm{d}R = \int_R p(X_b\vert X_R)p(X_a)p(R)\ \mathrm{d}R
\end{align}
However, the exact form of $p(X_b\vert X_R)$ is still unknown. We
approximate it with a similarity function $S(X_b, X_R)$, and denote
the resulting integral as $Z=\int_R S(X_b, X_R)p(R)\
\mathrm{d}R\approx \int_R p(X_b\vert X_R)p(R)\ \mathrm{d}R$.  We
define the similarity based on Gaussian similarity and possible
mappings $M$ of the vertices,\footnote{Because the vertices are
  connected in a way which forms a closed loop, we need only consider
  $2n$ mappings of the $n$ vertices (we assume uncertainty for which
  is the ``first'' vertex, and then which of its two neighbors is the
  ``second''). So, the possible orderings are of the form
  $M=\lbrace{}0, 1, \ldots{}, n\rbrace{}$, $M=\lbrace{}n, 0, \ldots{},
  n-1\rbrace{}$, etc.} i.e. $S(X_b,
X_R)=\frac{1}{2n}\sum_{M\in\mathbb{M}}\prod_{i=1}^n\mathcal{N}(X_b[i]\
\vert \ (MX_R)[i], \Sigma)$ where $i$ denotes the $i^{th}$ vertex. An
example stimulus and corresponding $S$ is shown in Fig.
\ref{fig:shapes}.

@@

We defined a \naive{} model which performs a hill-climbing search over
the similarity function until it reaches a (possibly local)
maximum. Once a maximum as been found, the model computes an estimate
of $Z$ by linearly interpolating between sampled rotations (e.g.,
Fig. \ref{fig:li}).

@@

We evaluated each model's performance on 20 randomly generated shapes
which had between three and six vertices, inclusive, (e.g., $X_a$ in
Fig.  \ref{fig:stimuli}). For each shape, we computed 18 ``same'' and
18 ``different'' stimuli pairs, with $R$ spaced at $20^\prime$
increments between 0 and 360, as in \cite{Shepard1971}. ``Same'' pairs
were created by rotating $X_a$ by $R$; the same was true for
``different'', except that $X_a$ was also reflected across the
$y$-axis. To gauge performance, we looked at \textit{response error
  rates}: how accurate was the model at choosing the correct
hypothesis? This was defined as the mean error ($\ME{}$), or fraction
of times the model chose incorrectly.  We additionally looked at
\textit{rotations}: for those ``same'' pairs which the model judged
correct, how correlated were the model's rotations with the true
angles of rotation?  We quantified this using the Pearson's
correlation coefficient $\rho$ for the true rotation ($R$) versus the
number of steps/rotations take by the model ($\vert
\mathbf{R}\vert$). Fig. \ref{fig:rotations} shows true rotations vs
number of steps/rotations by the model for individual stimuli
(\ref{fig:all-stimuli}) and average per true rotation
(\ref{fig:correct-stimuli}). This latter analysis can be qualitatively
compared to the results of \cite{Shepard1971} (Fig.
\ref{fig:mental-rotation}).

@@

The \naive{} model's response error rate was $\NaiveME{}$, which is
better than chance (equivalent to guessing randomly,
i.e. $\ME{}=0.5$). The correlation between the \naive{} model's
average rotation and the true angle of rotation was $\Naivecorr{}$
(Fig. \ref{fig:correct-stimuli}, left). For $R<\frac{\pi}{2}$, the
\naive{} model corresponds extremely well to the true angle of
rotation; this is because it needs to rotate less and is therefore
less likely to get stuck on local maxima. For $R>\frac{\pi}{2}$, we
see an increasing tendency to under-rotate due to getting stuck on
local maxima, as well as a tendency to over-rotate if the wrong
direction was initially chosen.

@@

The BQ model was much more accurate in choosing the correct hypothesis
than the other model ($\BQME{}$). The number of rotations computed by
the BQ model were strongly correlated with the true rotations
($\BQcorr$), a result which is qualitatively similar to that exhibited
by humans (Fig. \ref{fig:mental-rotation}
vs. Fig. \ref{fig:correct-stimuli}, right). Because the BQ model has
the capacity to ``reset'', it could recover from rotating in the
incorrect direction (e.g., Fig. \ref{fig:bq}) and thus did not
over-rotate as frequently.  It also under-rotated less frequently:
because the model continues to rotate until it is confident in its
estimate of $Z$, it does not get stuck as easily on local optima.

>>>>

Consider the images on the left in Fig.~\ref{fig:mental-rotation}. In
each panel, are the two depicted objects identical (except for a
rotation), or distinct? When presented with this mental rotation task,
people default to a strategy in which they visualize one object
rotating until it is congruent with the other
\cite{Shepard1971}. There is strong evidence for such mental
simulation: we can imagine three-dimensional objects in our minds and
manipulate them, to a certain extent, as if they were real
\cite{Kosslyn:2009tj}.  However, the use of mental simulation is
predicated on determining appropriate parameters to give the
simulation, analogous to determining exactly what computation should
be passed to a graphics processor.  In the case of the classic mental
rotation task, we might ask: How do people know which way to rotate
the object?  When should one stop rotating and accept the hypothesis
that the objects are different?

@@

We begin by first analyzing mental rotation at Marr's
\textit{computational} level \cite{Marr:1983to}: what is the problem
to be solved?  Formally, people are presented with two images, $X_a$
and $X_b$, which are the coordinates of the vertices of 2D shapes
(e.g., Fig.~\ref{fig:stimuli}). Participants must determine whether
$X_a$ and $X_b$ depict the same shape, i.e., whether $\exists
R\textrm{ s.t. } X_b=RX_a$, where $R$ is a rotation matrix. We can
formulate the judgment of whether $X_a$ and $X_b$ have the same
origins by deciding about two hypotheses, $h_0$: $\forall R\ X_b\neq
RX_a$ and $h_1$: $\exists R\textrm{ s.t. } X_b=RX_a$.  To compare the
hypotheses, we need to compute the posterior for each: $p(h\ \vert\
X_a, X_b)\propto p(X_a, X_b\ \vert\ h)p(h)$. Assuming the hypotheses
are equally likely \textit{a priori}, the prior term $p(h)$ will
cancel out when comparing $h_0$ and $h_1$, thus allowing us to focus
on the likelihoods, which are $p(X_a,\ X_b\ \vert \ h_0)=p(X_a)p(X_b)$
for $h_0$ and $p(X_a,\ X_b\ \vert \ h_1)=\int_R p(X_a) p(X_b\vert
X_a,R) p(R)\ \mathrm{d}R$ for $h_1$. From these likelihoods, we
compute the ratio $\ell$ which is given by $\ell=\left(\int_R p(X_b\
  \vert\ X_a, R)p(R)\ \mathrm{d}R\right) /\ p(X_b)$. If $\ell<1$, then
$h_0$ is the more likely hypothesis. If $\ell>1$, then $h_1$ is the
more likely hypothesis.

@@

We define the prior probability of shape $X$ to be
$p(X)=n!\left(\frac{1}{2\pi}\right)^n$ according to a generative
procedure.\footnote{A set of $n$ vertices could be chosen in any of
  $n!$ different ways, and each vertex is located at a random angle
  (between 0 and $2\pi$) and radius (between 0 and 1).} This gives us
the denominator of $\ell$. Computing the numerator is more difficult,
as we we cannot compute $p(X_b\vert X_a, R)$ directly. Instead, we
introduce a new variable $X_R$ denoting a mental image, which
approximates $RX_a$. The $X_R$ are generated by repeated application
of a function $\tau$, i.e. $X_R=RX_a=\tau(X_{R-r},
r)=\tau(\tau(X_{R-2r}, r), r)=\ldots{}=\tau^{(\frac{R}{r})}(X_a, r)$,
where $r$ is a small angle, and $\tau^{(i)}$ indicates $i$ recursive
applications of $\tau$. Using this sequential function, we get:
\begin{align}
  &p(X_a, X_b\ \vert \ h_1)=\int_R \int_{X} p(X_b\vert X) p(X\vert X_a, R)p(X_a)p(R)\ \mathrm{d}X\ \mathrm{d}R \nonumber \\
  &= \int_R \int_X p(X_b\vert X)\delta(\tau^{(\frac{R}{r})}(X_a, r)-X)p(X_a)p(R)\ \mathrm{d}X\ \mathrm{d}R = \int_R p(X_b\vert X_R)p(X_a)p(R)\ \mathrm{d}R
\end{align}
However, the exact form of $p(X_b\vert X_R)$ is still unknown. We
approximate it with a similarity function $S(X_b, X_R)$, and denote
the resulting integral as $Z=\int_R S(X_b, X_R)p(R)\
\mathrm{d}R\approx \int_R p(X_b\vert X_R)p(R)\ \mathrm{d}R$.  We
define the similarity based on Gaussian similarity and possible
mappings $M$ of the vertices,\footnote{Because the vertices are
  connected in a way which forms a closed loop, we need only consider
  $2n$ mappings of the $n$ vertices (we assume uncertainty for which
  is the ``first'' vertex, and then which of its two neighbors is the
  ``second''). So, the possible orderings are of the form
  $M=\lbrace{}0, 1, \ldots{}, n\rbrace{}$, $M=\lbrace{}n, 0, \ldots{},
  n-1\rbrace{}$, etc.} i.e. $S(X_b,
X_R)=\frac{1}{2n}\sum_{M\in\mathbb{M}}\prod_{i=1}^n\mathcal{N}(X_b[i]\
\vert \ (MX_R)[i], \Sigma)$ where $i$ denotes the $i^{th}$ vertex. An
example stimulus and corresponding $S$ is shown in
Fig.~\ref{fig:shapes}.

@@

We defined a \naive{} model which performs a hill-climbing search over
the similarity function until it reaches a (possibly local)
maximum. Once a maximum as been found, the model computes an estimate
of $Z$ by linearly interpolating between sampled rotations (e.g.,
Fig.~\ref{fig:li}).

@@

We evaluated each model's performance on 20 randomly generated shapes
which had between three and six vertices, inclusive, (e.g., $X_a$ in
Fig.~\ref{fig:stimuli}). For each shape, we computed 18 ``same'' and
18 ``different'' stimuli pairs, with $R$ spaced at $20^\prime$
increments between 0 and 360, as in \cite{Shepard1971}. ``Same'' pairs
were created by rotating $X_a$ by $R$; the same was true for
``different'', except that $X_a$ was also reflected across the
$y$-axis. To gauge performance, we looked at \textit{response error
  rates}: how accurate was the model at choosing the correct
hypothesis? This was defined as the mean error ($\ME{}$), or fraction
of times the model chose incorrectly.  We additionally looked at
\textit{rotations}: for those ``same'' pairs which the model judged
correct, how correlated were the model's rotations with the true
angles of rotation?  We quantified this using the Pearson's
correlation coefficient $\rho$ for the true rotation ($R$) versus the
number of steps/rotations take by the model ($\vert
\mathbf{R}\vert$). Fig.~\ref{fig:rotations} shows true rotations vs
number of steps/rotations by the model for individual stimuli
(\ref{fig:all-stimuli}) and average per true rotation
(\ref{fig:correct-stimuli}). This latter analysis can be qualitatively
compared to the results of \cite{Shepard1971}
(Fig.~\ref{fig:mental-rotation}).

@@

The \naive{} model's response error rate was $\NaiveME{}$, which is
better than chance (equivalent to guessing randomly,
i.e. $\ME{}=0.5$). The correlation between the \naive{} model's
average rotation and the true angle of rotation was $\Naivecorr{}$
(Fig.~\ref{fig:correct-stimuli}, left). For $R<\frac{\pi}{2}$, the
\naive{} model corresponds extremely well to the true angle of
rotation; this is because it needs to rotate less and is therefore
less likely to get stuck on local maxima. For $R>\frac{\pi}{2}$, we
see an increasing tendency to under-rotate due to getting stuck on
local maxima, as well as a tendency to over-rotate if the wrong
direction was initially chosen.

@@

The BQ model was much more accurate in choosing the correct hypothesis
than the other model ($\BQME{}$). The number of rotations computed by
the BQ model were strongly correlated with the true rotations
($\BQcorr$), a result which is qualitatively similar to that exhibited
by humans (Fig.~\ref{fig:mental-rotation}
vs. Fig.~\ref{fig:correct-stimuli}, right). Because the BQ model has
the capacity to ``reset'', it could recover from rotating in the
incorrect direction (e.g., Fig.~\ref{fig:bq}) and thus did not
over-rotate as frequently.  It also under-rotated less frequently:
because the model continues to rotate until it is confident in its
estimate of $Z$, it does not get stuck as easily on local optima.
